How fluid-mechanical erosion creates anisotropic porous media

### 2023-01-08 docs/PhysicsPaper/StokesErosion.tex
NM

\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\black}[1]{#1}
\newcommand{\edit}[1]{\red{#1}}

@@

% Maybe HOW fluid-mechanical erosion creates anisotropic porous media.
%\title{Fluid-mechanical erosion generates anisotropy in porous media}
%\title{Erosion leads to anisotropy in a porous medium}
%\title{The development of anisotropy in an eroding porous medium}

@@

\begin{abstract}
Using a Cauchy integral formulation of the boundary integral equations, we simulate the erosion of a porous medium comprised of up to 100 solid bodies embedded in a \edit{two-dimensional} Stokes flow. The grains of the medium are resolved individually and erode under the action of surface shear stress. Through nonlinear feedback with the surrounding flow fields, microscopic changes in grain morphology give way to larger-scale features in the medium such as channelization. The Cauchy-integral formulation and associated quadrature formulas enable us to resolve dense configurations of nearly contacting bodies.
We observe substantial anisotropy to develop over the course of erosion; that is, the configurations that result from erosion generally permit flow in the longitudinal direction more easily than in the transverse direction by up to a factor of six. These results suggest that the erosion of solid material from groundwater flows may contribute to previously observed anisotropy of natural porous media.
\end{abstract}
\maketitle

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% dematerialize, disintegrate
% internal, permeating, intervening

@@

%We employ the Stokes system for two reasons:~(1) this flow regime is appropriate for many groundwater applications \cite{bear1988dynamics}; and (2) the linearity of the Stokes system provides

@@

Consider an incompressible, Stokes flow inside a domain $\Omega$ containing $M$ erodable bodies. We take the outer boundary $\Gamma$ to be a slightly smoothed version of the boundary of $[-3,3] \times [-1,1]$. The fluid flow is primarily from left to right, so that the inlet and outlet are located at approximately $x=\pm 3$ (the actual locations are slightly curved versions of these vertical lines due to domain smoothing). The erodable bodies, with boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$, all sit inside of the central region $[-1,1] \times [-1,1]$. The boundary of the fluid domain is thus $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup \gamma_M$. The empty space to the left and right of $[-1,1] \times [-1,1]$ serve as buffer regions to allow the flow profile imposed at the inlet and outlet to gradually adjust to the presence of the bodies. The equations governing the velocity $\uu = (u,v)$ and pressure $p$ of the fluid consist of the incompressible, Stokes equations coupled to boundary conditions: \begin{equation}
\label{eqn:StokesEq}
  \begin{split}
    \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the erodable bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em prescribed outer wall velocity}.
  \end{split}
\end{equation}
\edit{The above equations are in dimensionless form, where space has been scaled on a domain length-scale, velocity has been scaled on a characteristic inlet value, and pressure has been scaled on the viscous force-per-unit-area that results from the choices of length and velocity scales.}
Meanwhile, $\UU$ represents the (dimensionless) fluid velocity imposed along the outer boundary $\Gamma$, in particular at the inlet and outlet, as well as along the top and bottom walls. In this work, we impose a {\em uniform} flow profile along $\Gamma$, i.e.~$\UU = (U,0)$, although other choices are possible, for example a Poiseuille profile as employed in previous work~\cite{chiu2020viscous, quaife2018boundary}. The advantages of the uniform profile are: (1) it will simplify the calculation of porous-medium properties, such as permeability and anisotropy, that will be described later; and (2), it may more realistically model the flow impinging upon a porous medium. We will allow the imposed flow speed to change with time $U = U(t)$ to enforce, for example, a desired pressure drop across the flow cell. 
%To nondimensionalize the above system, we set the fluid viscosity to unity, $\mu = 1$.

@@

While the Stokes equations~\eqref{eqn:StokesEq} provides a microscopic description of the detailed flow field $\uu$ penetrating the complex configuration of erodable bodies, a coarse-grained description can be obtained by treating the collection of bodies as a single porous-medium and homogenizing the flow-field through Darcy's law,
\begin{equation}
  \label{eqn:Darcy}
  \bq = - \frac{1}{\mu} \bvec{k} \grad p.
\end{equation}
\edit{Here, $p$ represents the pressure field and $\mu$ represents the fluid viscosity, which is set to unity due to the non-dimensionalization, $\mu=1$.}
Meanwhile, $\bq = (q_1, q_2)$ represents the {\em specific discharge}, which is the volume of water flowing through a unit cross sectional area of porous media per unit time; $\bq$ relates to the (interstitial) velocity $\uu$, by integrating $\uu$ over a sufficiently small control region and dividing by the total volume (including both fluid and solid) of the region. The parameter $\bvec{k}$ represents the {\em permeability} of the porous medium, which generally takes the form of a rank-2 tensor to permit different propensities to flow in different directions, i.e.~medium anisotropy.  
%The permeability may vary in space and will certainly vary in time as the bodies erode.

@@

%Since $\bq$ is a volume-average of the tracer velocity, $\qavg$ can be equivalently obtained by averaging the tracer velocity, which is how we will measure this quantity in the simulations.

@@

%verifying the relationship \eqref{perm_drag}, as will be done in Fig.~\ref{fig2}, validates a number of approximations made in the coarse-grained approach, such as (1) the homogenization of the flow field implicit in Darcy's law \eqref{eqn:Darcy}; (2) the calculation of the permeability from measurements of the flow rate and upstream/downstream pressures through \eqref{eqn:k11}.

@@

%
%Since all of the individual shapes are identical and isotropic (i.e.~circles), the resulting anisotropy is entirely due to the relative positions of the bodies, thus allowing us to define the {\em configurational} anisotropy as $\anis_C = k_{11}^{(c)} / k_{22}^{(c)}$. The total anisotropy, meanwhile, is the product of the shape and configurational anisotropy, $\anis = \anis_S \anis_C$, the latter of which we have measured. We can therefore deduce the shape anisotropy through $\anis_S = \anis/\anis_C$.

%-------------------------------------------------------------------%
\subsection{Tortuosity}
%
%We need to write this section.
%Could reference Souzy et al. JFM article on velocity distributions in porous media \cite{souzy2020velocity}

@@

% NOTE: I explain what ratio we compute in the results section, so we don't need to do it here.
%Then, another measure of the anisotropy is
%\begin{align}
 % \anistort = T_2/T_1.
%\end{align}
%\todo[inline]{Might eliminate this line so that we don't double-use anisotropy. In any case, I think $(T_2-1)/(T_1-1)$ might be more useful. Will give it thought.}
%As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

The previous section shows intriguing features to arise from the erosion of a single, random initialization of solid bodies. The observations naturally raise the question: are the trends specific to the particular simulation shown or do these features emerge robustly across different initial conditions? To answer this question, we now analyze the same quantities---permeability, tortuosity, and anisotropy---over an ensemble of simulations having different numbers of bodies and different random initializations. We analyze runs having $M$ = 20, 40, 60, 80, and 100 bodies, with at least three different instances of each. This data set represents roughly 2,100 hours of computational time.

>>>>

\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\black}[1]{#1}
\newcommand{\edit}[1]{\black{#1}}

@@

@@

\begin{abstract}
Using a Cauchy integral formulation of the boundary integral equations, we simulate the erosion of a porous medium comprised of up to 100 solid bodies embedded in a \edit{two-dimensional} Stokes flow. The grains of the medium are resolved individually and erode under the action of surface shear stress. Through nonlinear feedback with the surrounding flow fields, microscopic changes in grain morphology give way to larger-scale features in the medium such as channelization. The Cauchy-integral formulation and associated quadrature formulas enable us to resolve dense configurations of nearly contacting bodies.
We observe substantial anisotropy to develop over the course of erosion; that is, the configurations that result from erosion generally permit flow in the longitudinal direction more easily than in the transverse direction by up to a factor of six. These results suggest that the erosion of solid material from groundwater flows may contribute to previously observed anisotropy of natural porous media.
\end{abstract}
\maketitle
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

@@

Consider an incompressible, Stokes flow inside a domain $\Omega$ containing $M$ erodable bodies. We take the outer boundary $\Gamma$ to be a slightly smoothed version of the boundary of $[-3,3] \times [-1,1]$. The fluid flow is primarily from left to right, so that the inlet and outlet are located at approximately $x=\pm 3$ (the actual locations are slightly curved versions of these vertical lines due to domain smoothing). The erodable bodies, with boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$, all sit inside of the central region $[-1,1] \times [-1,1]$. The boundary of the fluid domain is thus $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup \gamma_M$. The empty space to the left and right of $[-1,1] \times [-1,1]$ serve as buffer regions to allow the flow profile imposed at the inlet and outlet to gradually adjust to the presence of the bodies. The equations governing the velocity $\uu = (u,v)$ and pressure $p$ of the fluid consist of the incompressible, Stokes equations coupled to boundary conditions: 
\begin{equation}
\label{eqn:StokesEq}
  \begin{split}
    \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the erodable bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em prescribed outer wall velocity}.
  \end{split}
\end{equation}
\edit{The above equations are in dimensionless form, where space has been scaled on a domain length-scale, velocity has been scaled on a characteristic inlet value, and pressure has been scaled on the viscous force-per-unit-area that results from the choices of length and velocity scales.}
Meanwhile, $\UU$ represents the (dimensionless) fluid velocity imposed along the outer boundary $\Gamma$, in particular at the inlet and outlet, as well as along the top and bottom walls. In this work, we impose a {\em uniform} flow profile along $\Gamma$, i.e.~$\UU = (U,0)$, although other choices are possible, for example a Poiseuille profile as employed in previous work~\cite{chiu2020viscous, quaife2018boundary}. The advantages of the uniform profile are: (1) it will simplify the calculation of porous-medium properties, such as permeability and anisotropy, that will be described later; and (2), it may more realistically model the flow impinging upon a porous medium. We will allow the imposed flow speed to change with time $U = U(t)$ to enforce, for example, a desired pressure drop across the flow cell.

@@

While the Stokes equations~\eqref{eqn:StokesEq} provides a microscopic description of the detailed flow field $\uu$ penetrating the complex configuration of erodable bodies, a coarse-grained description can be obtained by treating the collection of bodies as a single porous-medium and homogenizing the flow-field through Darcy's law,
\begin{equation}
  \label{eqn:Darcy}
  \bq = - \frac{1}{\mu} \bvec{k} \grad p.
\end{equation}
\edit{Here, $p$ represents the pressure field and $\mu$ represents the fluid viscosity, which is set to unity due to the non-dimensionalization, $\mu=1$.}
Meanwhile, $\bq = (q_1, q_2)$ represents the {\em specific discharge}, which is the volume of water flowing through a unit cross sectional area of porous media per unit time; $\bq$ relates to the (interstitial) velocity $\uu$, by integrating $\uu$ over a sufficiently small control region and dividing by the total volume (including both fluid and solid) of the region. The parameter $\bvec{k}$ represents the {\em permeability} of the porous medium, which generally takes the form of a rank-2 tensor to permit different propensities to flow in different directions, i.e.~medium anisotropy.

@@

@@

@@

%-------------------------------------------------------------------%
\subsection{Tortuosity}

@@

@@

The previous section shows intriguing features to arise from the erosion of a single, random initialization of solid bodies. The observations naturally raise the question: are the trends specific to the particular simulation shown or do these features emerge robustly across different initial conditions? To answer this question, we now analyze the same quantities---permeability, tortuosity, and anisotropy---over an ensemble of simulations having different numbers of bodies and different random initializations. We analyze runs having $M$ = 20, 40, 60, 80, and 100 bodies, with at least three different instances of each. \edit{We remark that varying $M$ permits us to study different initial permeabilities while keeping the initial porosity fixed at $\varphi = 0.4$.}
The entire data set represents roughly 2,100 hours of computational time.
